Data Specialization for a Staged Lambda Calculus

### 2014-11-04 paper/splitting.tex
Comments from marathon meeting.

We begin by describing how to split stage \bbtwo\ terms.

@@

Splitting a stage \bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
term $p$ and a \emph{resumer} function $l.r$, which contain all stage \bbone\
and \bbtwo\ subcomputations of $e$, respectively. Since stage \bbtwo\
computations can depend on stage \bbone, $r$ can depend on $p$ via the variable
$l$. The correctness of splitting thus requires that $e$ is equivalent to
$\letin l p r$.

Splitting, as implemented by the judgment $e \splittwosym [p,l.r]$
(\cref{fig:termSplitTwo}), is guided by the structure of the term $e$. 
For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
\end{lstlisting}
has two subterms, the latter of which itself has two subterms. Two are $\pause$s
containing stage \bbone\ computations, while the third has no stage \bbone\
content. The precomputation \emph{lifts} these computations in the same tree
structure: \verb|(1+2,(3+4,()))|.%
\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
The resumer is a function which computes the stage \bbtwo\ value of the original
term by mapping the results of the precomputation in place of the $\pause$s:
\verb|(L1,(L2,_)) => L1 < (L2 - 5)|.

In general, constants and variables have no stage \bbone\ content, so their
precomputations are \texttt{()}, and their resumers return the original term's
value. Non-staging constructs with $n$ subterms have precomputations which are
$n$-tuples of their subterms' precomputations, and resumers which pass each
component to the resumer of the respective subterm. (This behavior at
\texttt{case} and functions gives rise to speculative execution of
precomputations from underneath binders.) The $\prev$ and $\pause$ rules depends
on splitting stage \bbone\ terms, which is covered later.

\subsection{Partial Value Splitting}

>>>>

Since splitting stage \bbtwo\ terms is easier than splitting stage \bbone\ terms, 
we begin by describing how to split stage \bbtwo\ terms.

@@

Splitting a stage \bbtwo\ term $\coltwo e A$ produces a \emph{precomputation}
term $p$ and a \emph{resumer} function $l.r$, which contain all stage \bbone\
and \bbtwo\ subcomputations of $e$, respectively. Since stage \bbtwo\
computations can depend on stage \bbone\ computations, $r$ can depend on $p$ via the variable
$l$. The correctness of splitting thus requires that $e$ is equivalent to
$\letin l p r$.

% TODO this contract works because a stage-2 term reduces to a value
% so we can simply compare (p,l.r) to that value

Splitting stage \bbtwo\ terms, as implemented by the judgment $e \splittwosym [p,l.r]$
(\cref{fig:termSplitTwo}), is guided by the structure of the term $e$. 
For example, the stage \bbtwo\ term
\begin{lstlisting}
2`hold{`1` 1+2 `2`} < (hold{`1` 3+4 `2`} - 5)`
\end{lstlisting}
has two subterms, the latter of which itself has two subterms. Two are $\pause$s
containing stage \bbone\ computations, while the third has no stage \bbone\
content. The precomputation \emph{lifts} these computations in the same tree
structure: \verb|(1+2,(3+4,()))|.%
\footnote{In later examples, we will optimize out unnecessary \texttt{()}s.}
The resumer is a function which computes the stage \bbtwo\ value of the original
term by mapping the results of the precomputation in place of the $\pause$s:
\verb|(L1,(L2,_)) => L1 < (L2 - 5)|.

In general, constants and variables have no stage \bbone\ content, so their
precomputations are \texttt{()}, and their resumers return the original term's
value. Non-staging constructs with $n$ subterms have precomputations which are
$n$-tuples of their subterms' precomputations, and resumers which pass each
component to the resumer of the respective subterm. (This behavior at
\texttt{case} and functions causes precomputations to be lifted out from 
underneath stage \bbtwo\ binders.) The $\prev$ and $\pause$ rules depends
on splitting stage \bbone\ terms, which is covered later.

\TODO
The contract for splitting stage \bbone\ terms is more complex, because stage \bbone\ 
terms reduce to a pair of a future context and a partial value, not simply to a value.
